---
title: "Modified-Poisson-Regresion"
author: "Diego Pajarito"
date: "2026-08-27"
output: html_document
---




---
title: "Association Between Urban Informality and Maternal Mortality"
author: "Diego Pajarito Grajales"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

# Introduction

This analysis evaluates whether mortality rates differ across grid cells classified as having low, medium, or high urban informality. Population counts are used as an offset so that estimated effects represent differences in mortality rates rather than differences in population size.

Models:

a) Poisson
b) Negative Binomial


```{r setup, message=FALSE, warning=FALSE, echo=FALSE}
library(sf)
library(dplyr)
library(jsonlite)
library(MASS)
library(broom)
library(ggplot2)
library(knitr)
library(kableExtra)
library(performance)
library(DHARMa)
library(tidyr)
library(leaflet)
```

# Data 

##  Inputs:

Data inputs are the model results from IDEAMAPS (deprivation), mortality cases from
Kassep and population (WorlPOP women of childbearing age.

```{r}
mortality_json <- "./data/maternaldeaths.geojson"
grid_gpkg <- "./data/model-output-classification.gpkg"
grid_layer <- "deprivation-classification"
pop_gpkg <- "./data/pop-grid-kano-centroids.gpkg"
pop_layer <- "pop-grid-kano-centroids"
```

## Load Mortality Cases

This is a JSON file recording maternal mortality cases from KASSEP

```{r read-mortality-cases, echo=FALSE}

mortality_sf <- st_read(mortality_json, quiet = TRUE)
cat("Cases loaded. Rows:", nrow(mortality_sf), "\n")

# head(mortality_sf)
```

## Load EmOC Deprivation Grid

This is GeoPackage has the cells classified by EmOC access deprivation.
There are two classifications applied to the EmOC access deprivation:

One including the three categories and a second one where Medium and High 
are conflated to reduce the classification to two categories.


```{r read-emoc-deprivation, echo=FALSE}
grid <- st_read(grid_gpkg, layer = grid_layer, quiet = TRUE)

grid <- grid %>%
  mutate(
    informality_class = case_when(
      result == 0 ~ "Low",
      result == 1 ~ "Medium",
      result == 2 ~ "High",
      TRUE ~ NA_character_
    ),
    informality_class = factor(
      informality_class,
      levels = c("Low", "Medium", "High")
    ),
    informality_class_mh = case_when(
      result == 0 ~ "Low",
      result == 1 ~ "Medium-or-High",
      result == 2 ~ "Medium-or-High",
      TRUE ~ NA_character_
    ),
    informality_class_mh = factor(
      informality_class_mh,
      levels = c("Low", "Medium-or-High")
    )
  )

cat("Deprivation loaded. Cells:", nrow(grid), "\n")

```

## Load women of childbearing age population to the Grid

This Geopackage has cells with the estimate number of women of childbearing age.
Such values are the population values used in the model. The values are transferred
by applying and spatial join operation. Since the final values are the result of 
estimates based on the presence of building footprints, the following columns are
used:

- pop: estimated number of women of childbearing age in the cell
- pop_grid_pop: estimated population from the 1km x 1km grid cell (WorlPOP)
- bcount: number of building footprints used to distribute population (1km -> 100m)

```{r read-population-childbearingage, echo=FALSE}
popgrid <- st_read(pop_gpkg, layer = pop_layer, quiet = TRUE)

pop_centroids <- popgrid %>%
  dplyr::select(pop, pop_grid_pop, bcount)

grid <- grid %>%
  st_join(
    pop_centroids,
    join = st_contains
  )

cat("Population loaded. Cells:", nrow(popgrid), "\n")
#summary(popgrid)
```



# Data Preparation

Preparation tasks include operations such as
 - Project Layers to Common CRS

```{r project-df-crs, echo=FALSE}
mortality_sf <- st_transform(mortality_sf, st_crs(grid))
popgrid <- st_transform(popgrid, st_crs(grid))
```

 - Assign and count Mortality Cases to Grid Cells

```{r count-mortality-cases, echo=FALSE}
case_grid <- st_join(
  mortality_sf,
  grid,
  join = st_within,
  left = FALSE
)

case_counts <- case_grid %>%
  st_drop_geometry() %>%
  count(grid_id, name = "deaths")
```

## Analysis data set

To prepare the analysis data set, it was required to consider that the overall aim 
is to compare how mortality rates (mortality cases within the population) change, 
or not change, accross the different EmOC acces deprivation categories.

For this, the dataset should i) exclude gridcells where there's no population of
women of child bearing age and ii) consider areas with and withouth recorded
cases of maternal mortality


```{r create-analysis-dataset}
analysis_df <- grid %>%
  dplyr::filter(pop>0) %>% 
  left_join(case_counts, by = "grid_id") %>% 
  mutate(
    deaths = ifelse(is.na(deaths), 0, deaths)
  ) %>%
  st_drop_geometry() %>%
  dplyr::select(Accessibility_Index_Standard, informality_class, informality_class_mh, pop, deaths)

summary(analysis_df)
```


### Subset B: grid cells only with recorded mortality cases

This is a secondary dataset used for deeper analysis

```{r subA-create-dataset, echo=FALSE}
analysis_df_b <- grid %>%
  left_join(case_counts, by = "grid_id") %>% 
  mutate(
    deaths = ifelse(is.na(deaths), 0, deaths)
  ) %>%
  st_drop_geometry() %>%
  dplyr::filter(deaths>0) %>% 
  dplyr::select(Accessibility_Index_Standard, informality_class, informality_class_mh, pop, deaths)

summary(analysis_df_b)
```


# Model fitting

## Step 1: Poisson Model without adjusting by population

### Model fit for the three deprivation categories

This model serves as an initial exploration of the possible relationship between
EmOC access deprivation and the count of mortality cases. However, the results 
are not optimal due to the lack of adjustment by population.

One of the drawbacks of not accounting for population size is that larger cells 
may naturally have more deaths simply because more people live there.

This model assumes H0: Beta = 0, Or equivalent to IRR = 0
A key to interpret the model is the IRR (effect size) and the confidence interval.

Below an example of how this could be interpreted: 

IRR = 1.5 for Medium → medium informality cells have 50% more deaths than low informality cells.
IRR = 2.0 for High → high informality cells have twice as many deaths as low informality cells.


```{r model-subA-unadjusted}

poisson_unadjusted <- glm(
  deaths ~ informality_class,
  family = poisson(link = "log"),
  data = analysis_df
)
#summary(poisson_unadjusted)

# Extract coefficients
results <- tidy(
 poisson_unadjusted,
 exponentiate = TRUE,
 conf.int = TRUE,
)
# Keep only deprivation effects
results <- results %>%
  filter(term != "(Intercept)") %>%
  mutate(
  term = gsub("informality_class", "", term)
)

results

```
### Model fit for deprivation categories conlfated to two


```{r model-subA-unadjusted-conflated}

poisson_unadjusted_conflated <- glm(
  deaths ~ informality_class_mh,
  family = poisson(link = "log"),
  data = analysis_df_b
)
#summary(poisson_unadjusted_conflated)

# Extract coefficients
results <- tidy(
 poisson_unadjusted_conflated,
 exponentiate = TRUE,
 conf.int = TRUE,
)
# Keep only deprivation effects
results <- results %>%
  filter(term != "(Intercept)") %>%
  mutate(
  term = gsub("informality_class_mh", "", term)
)

results

```





## Step 2: Negative Binomial Model


The negative binomial model includes control by the population, in this case, 
of women of childbearing age. This compares the mortality rates between grid cells
in the different categories of EmOC access deprivation. This model allows to 
describe how much higher (or lower) is mortality in other categories compared with
low-deprivation cells.

The model considers facts related to maternal health such as death cases being
rare events, unevenly distributed and often clustered spatially that leave many
grid cells with ceros.

This model assumes H0: Beta = 0, Or equivalent to IRR = 0
A key to interpret the model is the IRR (effect size) and the 95% confidence interval.

### Model fit for the three deprivation categories


```{r model-negativebinomial}

nb_model <- glm.nb(
 deaths ~ informality_class + offset(log(pop)),
 data = analysis_df
)

summary(nb_model)
exp(cbind(
  IRR = coef(nb_model),
  confint(nb_model)
))

# Extract coefficients
results <- tidy(
 nb_model,
 conf.int = TRUE,
 exponentiate = TRUE
)

# Keep only deprivation effects
results <- results %>%
  filter(term != "(Intercept)") %>%
  mutate(
  term = gsub("informality_class", "", term)
)

results
```


### Model fit for the deprivation categories conflated to two 

```{r model-negativebinomial_conflated}

nb_model_mh <- glm.nb(
 deaths ~ informality_class_mh + offset(log(pop)),
 data = analysis_df
)

summary(nb_model_mh)
exp(cbind(
  IRR = coef(nb_model_mh),
  confint(nb_model_mh)
))

# Extract coefficients
results <- tidy(
 nb_model_mh,
 conf.int = TRUE,
 exponentiate = TRUE
)

# Keep only deprivation effects
results <- results %>%
  filter(term != "(Intercept)") %>%
  mutate(
  term = gsub("informality_class_mh", "", term)
)

results
```

### Model fit for only grid cells with maternal death cases


```{r model-negativebinomial_onlycasecells}

nb_model_case <- glm.nb(
 deaths ~ informality_class + offset(log(pop)),
 data = analysis_df_b
)
summary(nb_model_case)
exp(cbind(
  IRR = coef(nb_model_case),
  confint(nb_model_case)
))

# Extract coefficients
results <- tidy(
 nb_model_mh,
 conf.int = TRUE,
 exponentiate = TRUE
)

# Keep only deprivation effects
results <- results %>%
  filter(term != "(Intercept)") %>%
  mutate(
  term = gsub("informality_class", "", term)
)

results
```



## Step 3: Comparing Binomial GLM vs Poisson GLM 

There are some factors that serve to decide which models are more adequate for the
analysis. The comparison uses overdispertion, AIC and


### Estimating a Poisson GLM model controlled by population

```{r model-poison-coeficients}
pois_model <- glm(
  deaths ~ informality_class +
  offset(log(pop)),
  family = poisson(),
  data = analysis_df
)

# Extract coefficients
results <- tidy(
 pois_model,
 exponentiate = TRUE,
 conf.int = TRUE,
)
# Keep only deprivation effects
results <- results %>%
  filter(term != "(Intercept)") %>%
  mutate(
  term = gsub("informality_class_mh", "", term)
)

results
```


### Over dispersion

Estimating 

deviance(poisson_model) / df.residual(poisson_model)

sum(residuals(poisson_model, type="pearson")^2) /
  df.residual(poisson_model)


Ratio Value     Interpretation
 - aprox 1      Poisson assumption reasonable
 - 1.5 - 2      Mild overdispersion
 -    > 2       Substantial overdispersion
 -    > 4       Strong overdispersion
 
 for mortality data in grids, values above 2 are very common


```{r model-comparison-overdispersion}


rat_disp_v1 <- deviance(pois_model) / df.residual(pois_model)

rat_disp_v2 <- sum(residuals(pois_model, type="pearson")^2) /
  df.residual(pois_model)

data.frame(c("Deviance/Residual ", "sum(res--pearson)^2/residual "), c(rat_disp_v1, rat_disp_v2))


```

### AIC Values comparison 

For this comparison, the two models will provide AIC values with the lower AIC values
indicates the preferred model. The following are indications to support the 
decision 

  ΔAIC          Interpretation
 -   < 2        Models essentially equivalent
 - 2 - 4        Weak support for lower AIC model
 - 4 - 10       Moderate support
 -   > 10       Strong support



```{r model-comparison-AIC}



AIC(pois_model, nb_model)

```




## Step 3: Aggregating grid cells into 1km grid size


Having 100m grid cells makes the case of having thousands of cero values for 
mortality. Looking for an option to reduce this proportion, we tried to aggregate 
the cells. Since the model uses the 1km grid from WorldPOP 

The process group the cells based on the unique value for the estimated population
"pop_grid_pop", dissolves the geometries and estimate new values for the
accessibility index (avg), building count, and deprivation category (mode)


```{r aggregating-grid-1k}

get_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

grid1k <- grid %>%
  group_by(pop_grid_pop) %>%
  summarise(
    bcount = sum(bcount, na.rm = TRUE),
    grid_count = n(),
    Accessibility_Index_avg = mean(
      Accessibility_Index_Standard,
      na.rm = TRUE
    ),
    informality_mode = get_mode(result),
    .groups = "drop"
  ) %>%
  rename(pop = pop_grid_pop)

grid1k <- tibble::rowid_to_column(grid1k, "id")

```

### Estimating the new accessibility index

The index is estimated in two ways, one is the use of established thresholds
to the newly averaged index and the second comes form the mode value.

# Classify the accessibility into three categories based on the standardized Accessibility Index: 0 for High accessibility, 1 for Medium accessibility, and 2 for Low accessibility. The thresholds for classification are set at 0.005 and 0.02, which can be adjusted based on the distribution of the Accessibility Index in the dataset.

The thresholds used for the 100m grid cell are applied to the aggregated cells as:
High: Value 2. Accessibility index (0.02 - 1)
Medium: Value 1. Accessibility index (0.005 - 0.02)
Low: Value 0. Accessibility index (0 - 0.005)


```{r classifying-deprivation-1k, echo=FALSE}

grid1k <- grid1k %>%
  mutate(
    informality_class_mode = case_when(
      informality_mode == 0 ~ "Low",
      informality_mode == 1 ~ "Medium",
      informality_mode == 2 ~ "High",
      TRUE ~ NA_character_
    ),
    informality_class_mode = factor(
      informality_class_mode,
      levels = c("Low", "Medium", "High")
    ),
    informality_class = case_when(
      Accessibility_Index_avg >= 0 & Accessibility_Index_avg <= 0.005  ~ "High",
      Accessibility_Index_avg > 0.005 & Accessibility_Index_avg <= 0.02 ~ "Medium",
      Accessibility_Index_avg > 0.02 & Accessibility_Index_avg <= 1 ~ "Low",
      TRUE ~ NA_character_
    ),
    informality_class = factor(
      informality_class,
      levels = c("Low", "Medium", "High")
    )
  )

cat("Deprivation processed Cells:", nrow(grid1k), "\n")

```

### Counting mortality cases


```{r count-mortality-cases-1k, echo=FALSE}
case_grid_1k <- st_join(
  mortality_sf,
  grid1k,
  join = st_within,
  left = FALSE
)

case_counts_1k <- case_grid_1k %>%
  st_drop_geometry() %>%
  count(id, name = "deaths")
```




### Analysis data set that includes maternal death cases count


```{r create-analysis-dataset-1k}
analysis_1k <- grid1k %>%
  left_join(case_counts_1k, by = "id") %>% 
  mutate(
    deaths = ifelse(is.na(deaths), 0, deaths)
  ) %>%
  st_drop_geometry() %>%
  dplyr::select(Accessibility_Index_avg, informality_class, informality_mode, pop, deaths)

summary(analysis_1k)
```


### Negative Binomial Model using the 1k grid


The negative binomial model includes control by the population, in this case, 
of women of childbearing age. This compares the mortality rates between grid cells
in the different categories of EmOC access deprivation. This model allows to 
describe how much higher (or lower) is mortality in other categories compared with
low-deprivation cells.

The model considers facts related to maternal health such as death cases being
rare events, unevenly distributed and often clustered spatially that leave many
grid cells with ceros.

This model assumes H0: Beta = 0, Or equivalent to IRR = 0
A key to interpret the model is the IRR (effect size) and the 95% confidence interval.

### Model fit for the three deprivation categories


```{r model-negativebinomial-1k}

nb_model_1k <- glm.nb(
 deaths ~ informality_class + offset(log(pop)),
 data = analysis_1k
)

#summary(nb_model)
exp(cbind(
  IRR = coef(nb_model_1k),
  confint(nb_model_1k)
))

# Extract coefficients
results <- tidy(
 nb_model_1k,
 conf.int = TRUE,
 exponentiate = TRUE
)

# Keep only deprivation effects
results <- results %>%
  filter(term != "(Intercept)") %>%
  mutate(
  term = gsub("informality_class", "", term)
)

results
```



```{r aggregated-grid-plot}

pal <- colorFactor(
  palette = c("blue", "orange", "red"),
  domain = grid1k$informality_class
)

leaflet(grid1k) %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  addPolygons(
    fillColor = ~pal(informality_class),
    fillOpacity = 0.7,
    color = "black",
    weight = 0.5,
    popup = ~paste(
      "<b>Category:</b>", informality_class, "<br>",
      "<b>Category Mode:</b>", informality_mode, "<br>",
      "<b>Population:</b>", pop, "<br>",
      "<b>Bcount:</b>", bcount, "<br>",
      "<b>Accessibility Index:</b>",
      round(Accessibility_Index_avg, 4)
    )
  ) %>%
  addCircleMarkers(
    data = mortality_sf,
    radius = 5,
    color = "black",fillColor = "white",
    fillOpacity = 0.9,
    stroke = TRUE,
    weight = 1
  ) %>%
  addLegend(
    "bottomright",
    pal = pal,
    values = ~informality_class,
    title = "EmOC Accessibility"
  )

```


