---
title: "Estimating Risk Surface"
output:
  html_document:
    df_print: paged
---

# Preliminary tasks 

## Preparation
This notebook estimates a spatial risk surface for maternal deaths in Kano, Nigeria,
based on the kernel density ratio method: https://arxiv.org/pdf/1707.06888

Key optimisations applied:
- WorldPop raster is aggregated before point generation to reduce memory load
- Population points are capped at a maximum sample size (10,000)
- Fixed bandwidth used first for speed; adaptive bandwidth optional
- Duplicate chunk names fixed
- Cases/Controls variable naming corrected
- All data reprojected to Mollweide (ESRI:54009) for equal-area analysis




### Install required packages

There is a selection of packages required for the risk surface analysys. Among them, geospatial data management packages and "sparr".
```{r install-packages, include=FALSE, eval=FALSE}
# Run once, then set eval=FALSE
install.packages(c(
  "sparr", "terra", "sf", "spatstat", "dplyr", "ggplot2",
  "tidyterra", "ggspatial", "viridis", "cowplot", "rlang",
  "spdep", "rgeoda", "dunn.test", "scico"
))
```



### Load required packages
```{r load-packages, include=FALSE}
library(sf)
library(sparr)
library(terra)       # Modern replacement for raster package
library(spatstat)
library(dplyr)
library(ggplot2)
library(tidyterra)
library(ggspatial)
library(cowplot)
library(viridis)
library(spdep)     # Added to estimate local Moran's and bivariate local Moran's
library(rgeoda)
library(dunn.test)
library(scico)
```
---

# Load Data

## Model outputs
This data set contains the outputs for the "Emergency Obstetric Care Access Deprivation" model generated by the IDEAMAPS data ecosystem.
```{r read-model-outputs, include=FALSE}
data <- read.csv("data/model-output-classification.csv")
head(data)
```

## Model outputs
load geopackage
```{r model-outputs, include=FALSE}
model_outputs <- sf::st_read("data/model-output-classification.gpkg", quiet = TRUE)
head(model_outputs)
```

## Maternal deaths (Cases)
This data, provided by the KASSEP project, compiles cases of maternal deaths in Kano
```{r read-maternal-deaths, include=FALSE}
maternal_deaths <- sf::st_read("data/maternaldeaths.geojson", quiet = TRUE)
head(maternal_deaths)
```

## Study area boundary
```{r read-study-area, include=FALSE}
kano <- sf::st_read("data/grid-boundary-kano.gpkg", quiet = TRUE)
head(kano)
```

## Population data (Women of childbearing age 15-49, 100m resolution)
```{r read-population, include=FALSE}
population <- sf::st_read("data/pop-grid-kano-centroids.gpkg", quiet = TRUE)
head(population)
```

---

# Main Analysis

## Step 1 — Reproject to Mollweide

Equal-area projection required by `spatstat`; all layers must share the same CRS.

```{r reproject, include=FALSE}
CRS_MOLLWEIDE <- "ESRI:54009"

kano_proj            <- sf::st_transform(kano,            crs = CRS_MOLLWEIDE)
maternal_deaths_proj <- sf::st_transform(maternal_deaths, crs = CRS_MOLLWEIDE)
population_proj      <- sf::st_transform(population,      crs = CRS_MOLLWEIDE)
model_proj           <- sf::st_transform(model_outputs,   crs = CRS_MOLLWEIDE)
```

## Step 2 — Clip deaths to 10 km study-area buffer

Deaths outside the study boundary have no matching population controls
and would inflate the risk surface.

```{r clip-deaths}
kano_buffer         <- sf::st_buffer(kano_proj, dist = 10000)
maternal_deaths_kano <- sf::st_filter(maternal_deaths_proj, kano_buffer)

cat("Deaths retained:", nrow(maternal_deaths_kano),
    "of", nrow(maternal_deaths_proj), "\n")

# Quick visual check
plot(sf::st_geometry(kano_buffer),
     main = "Maternal Deaths — 10 km Buffer", border = "grey50", lty = 2)
plot(sf::st_geometry(kano_proj),        add = TRUE, border = "black")
plot(sf::st_geometry(maternal_deaths_kano), add = TRUE, col = "red", pch = 16, cex = 0.8)
```
# Layer 1 — Where do the deaths actually fall?
```{r deaths-by-deprivation}
# ---- Spatial join: each death point → its underlying model grid cell ----
# st_join with st_intersects assigns each death the deprivation class
# of the grid cell it falls in.
deaths_with_class <- sf::st_join(
  maternal_deaths_kano,
  model_proj["result"],
  join = sf::st_intersects
)

# Some deaths may fall in the 10 km buffer but outside the actual
# model grid — drop them so we compare like with like.
deaths_in_grid <- deaths_with_class |> dplyr::filter(!is.na(result))

cat("Deaths inside model grid:", nrow(deaths_in_grid),
    "of", nrow(maternal_deaths_kano), "\n\n")

# ---- Observed distribution: where deaths actually fell ----
obs <- table(factor(deaths_in_grid$result,
                    levels = 0:2,
                    labels = c("Low", "Medium", "High")))

# ---- Expected distribution: where women of childbearing age live ----
# This is the NULL hypothesis: deaths track population, not deprivation.
# If the chi-squared test rejects this null, we have evidence that
# deprivation matters beyond population density.
pop_with_class <- sf::st_join(
  womenrepage,
  model_proj["result"],
  join = sf::st_intersects
) |> dplyr::filter(!is.na(result))

exp_pct <- prop.table(table(factor(pop_with_class$result,
                                    levels = 0:2,
                                    labels = c("Low", "Medium", "High"))))
exp <- as.numeric(exp_pct) * sum(obs)

# ---- Chi-squared goodness-of-fit test ----
# H0: deaths distributed proportionally to population
# H1: deaths concentrated differently than population (i.e. deprivation
#     adds explanatory power beyond population density)
chi_test <- chisq.test(x = as.numeric(obs), p = as.numeric(exp_pct))

# ---- Paper-ready output ----
cat("Deprivation class | Observed deaths | Expected (pop-weighted)\n")
cat("------------------+-----------------+------------------------\n")
for (i in seq_along(obs)) {
  cat(sprintf("%-17s | %15d | %22.1f\n",
              names(obs)[i], obs[i], exp[i]))
}
cat(sprintf("\nχ² = %.2f, df = %d, p = %.3g\n",
            chi_test$statistic, chi_test$parameter, chi_test$p.value))

cat(sprintf("\n%.1f%% of deaths occurred in Medium or High deprivation areas\n",
            100 * (obs["Medium"] + obs["High"]) / sum(obs)))
cat(sprintf("(vs %.1f%% expected from population distribution alone)\n",
            100 * (exp_pct["Medium"] + exp_pct["High"])))
```

## Step 3: Generate population points from 100m grid polygons

Each polygon represents a 100m grid cell with a population value.
Random points are generated within each polygon proportional to population.
SCALE_FACTOR reduces total point count while preserving relative spatial distribution.
st_sample() places points randomly inside each polygon - no manual cell size needed.

Points are placed randomly inside each 100 m grid cell, proportional to population
(`SCALE_FACTOR = 10` reduces total count while preserving spatial distribution).

```{r generate-pop-points}
pop_valid <- population_proj |>
  dplyr::filter(!is.na(pop), pop > 0) |>
  dplyr::mutate(n_points = round(pop / 10)) |>
  dplyr::filter(n_points > 0)

cat("Cells used:", nrow(pop_valid),
    "| Points to generate:", sum(pop_valid$n_points), "\n")

set.seed(42)
points_list <- lapply(seq_len(nrow(pop_valid)), function(i) {
  n    <- pop_valid$n_points[i]
  bbox <- sf::st_bbox(pop_valid[i, ])
  data.frame(
    x = runif(n, bbox["xmin"], bbox["xmax"]),
    y = runif(n, bbox["ymin"], bbox["ymax"])
  )
})

womenrepage <- sf::st_as_sf(
  do.call(rbind, points_list),
  coords = c("x", "y"),
  crs    = sf::st_crs(population_proj)
)

cat("Population points generated:", nrow(womenrepage), "\n")
```



## Step 4: Build spatial window from Kano polygon boundary

Using the actual Kano boundary polygon (not a bounding box rectangle)
ensures the risk surface matches the study area shape exactly.
This requires projected coordinates - hence the UTM re-projection above.

```{r build-window, include=FALSE}
kano_window <- spatstat.geom::as.owin(sf::st_cast(kano_buffer, "POLYGON"))
```


## Step 5: Create PPP objects

PPP objects are two
- Cases: Maternal dates within Kano study area
- Controls: population at risk defined as women of childbearing age

```{r create-cases-ppp, include=FALSE}
# Cases: maternal deaths within Kano study area
deaths_coords <- sf::st_coordinates(maternal_deaths_kano)
deaths_ppp    <- spatstat.geom::ppp(
  x      = deaths_coords[, 1],
  y      = deaths_coords[, 2],
  window = kano_window
)
```

```{r create-controls-ppp, include=FALSE}
# Controls: population at risk (women of childbearing age)
pop_coords <- sf::st_coordinates(womenrepage)
pop_ppp    <- spatstat.geom::ppp(
  x      = pop_coords[, 1],
  y      = pop_coords[, 2],
  window = kano_window
)
```

## Step 6: Estimate risk surface

### Option A: Fixed bandwidth (run first to verify everything works)
```{r risk-fixed, include=TRUE}
rr_fixed <- risk(f = deaths_ppp, g = pop_ppp, doplot = TRUE)
```

### Option B: Geometric bandwidth (recommended)
```{r risk-geometric, include=TRUE}
h0 <- OS(deaths_ppp, nstar = "geometric")
cat("Optimal bandwidth:", h0, "\n")

rr_geom <- risk(
  f        = deaths_ppp,
  g        = pop_ppp,
  h0       = h0,
  tolerate = TRUE,
  doplot   = TRUE
)
```

### Option C: Adaptive bandwidth (most accurate, slowest)
# Set eval=TRUE in chunk header when ready to run - may take 10-30 mins
```{r risk-adaptive, eval=TRUE}
hpilot.f <- BOOT.density(deaths_ppp)
hpilot.g <- BOOT.density(pop_ppp)

rr_adapt <- risk(
  f               = deaths_ppp,
  g               = pop_ppp,
  h0              = h0,
  adapt           = TRUE,
  tolerate        = TRUE,
  hp              = c(hpilot.f, hpilot.g),
  pilot.symmetry  = "pooled",
  davies.baddeley = 0.05
)

plot(rr_adapt, main = "Risk Surface - Adaptive Bandwidth")
```

## Step 7 — Map risk surface (ggplot)

```{r plot-risk-surface}
# Convert to terra rasters and set CRS
rr_raster <- terra::rast(rr_adapt$rr); terra::crs(rr_raster) <- "ESRI:54009"
rr_p      <- terra::rast(rr_adapt$P);  terra::crs(rr_p)      <- "ESRI:54009"

# Crop and mask to study area
rr_raster <- terra::mask(terra::crop(rr_raster, terra::vect(kano_proj)), terra::vect(kano_proj))
rr_p      <- terra::mask(terra::crop(rr_p,      terra::vect(kano_proj)), terra::vect(kano_proj))

kano_sf <- sf::st_as_sf(kano_proj)

p_risk <- ggplot() +
  tidyterra::geom_spatraster(data = rr_raster) +
  scale_fill_gradientn(
    colours  = c("#000080", "#800080", "#FF0000", "#FF4500", "#FF8000", "#FFFF00"),
    na.value = "transparent",
    name     = "Log Risk\nRatio"
  ) +
  geom_sf(data = kano_sf, fill = NA, color = "black", linewidth = 0.5) +
  tidyterra::geom_spatraster_contour(
    data = rr_p, breaks = 0.05, color = "black", linewidth = 0.6
  ) +
  geom_sf(data = maternal_deaths_kano,
          color = "black", fill = "yellow", shape = 21, size = 1.5, stroke = 0.4) +
  ggspatial::annotation_scale(location = "bl", width_hint = 0.2, text_cex = 0.7) +
  ggspatial::annotation_north_arrow(
    location    = "br", which_north = "true",
    pad_x       = unit(0.1, "in"), pad_y = unit(0.1, "in"),
    style       = north_arrow_fancy_orienteering(text_size = 6)
  ) +
  labs(
    title    = "Maternal Death Risk Surface — Kano, Nigeria",
    subtitle = "Kernel density ratio (adaptive bandwidth) | Contours: p < 0.05"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle = element_text(size = 9, color = "grey40", hjust = 0.5),
    legend.position = "right"
  )

print(p_risk)
```

```{r plot-risk-surface-paper-style, fig.width=11, fig.height=8}
rr_raster <- terra::rast(rr_adapt$rr)
terra::crs(rr_raster) <- "ESRI:54009"

rr_p <- terra::rast(rr_adapt$P)
terra::crs(rr_p) <- "ESRI:54009"

rr_raster <- terra::mask(
  terra::crop(rr_raster, terra::vect(kano_proj)),
  terra::vect(kano_proj)
)

rr_p <- terra::mask(
  terra::crop(rr_p, terra::vect(kano_proj)),
  terra::vect(kano_proj)
)
rr_raster_smooth <- terra::disagg(rr_raster, fact = 6)
rr_p_smooth      <- terra::disagg(rr_p, fact = 6)

kano_sf <- sf::st_as_sf(kano_proj)

p_risk <- ggplot() +

  tidyterra::geom_spatraster(
    data = rr_raster_smooth,
    interpolate = TRUE,
    alpha = 0.97
  ) +

  scale_fill_distiller(
    palette   = "RdBu",
    direction = -1,
    limits    = c(-1.5, 1.5),
    na.value  = "transparent",
    name      = "Log Risk\nRatio",

    guide = guide_colorbar(
      order = 1,
      barheight = unit(6, "cm"),
      barwidth  = unit(0.7, "cm"),
      title.position = "top",
      title.hjust = 0.5
    )
  ) +

  tidyterra::geom_spatraster_contour(
    data = rr_p_smooth,
    aes(linetype = "Statistically significant\n(p < 0.05)"),
    breaks = 0.05,
    color = "black",
    linewidth = 0.35,
    alpha = 0.85
  ) +

  geom_sf(
    data = kano_sf,
    aes(linetype = "Kano State boundary"),
    fill = NA,
    color = "black",
    linewidth = 0.55
  ) +

  geom_sf(
    data = maternal_deaths_kano,
    aes(shape = "Maternal death case location"),
    fill = "#F4E409",
    color = "black",
    size = 1.8,
    stroke = 0.35,
    alpha = 0.95
  ) +

  scale_shape_manual(
    values = c(
      "Maternal death case location" = 21
    ),

    name = NULL,

    guide = guide_legend(
      order = 2,
      override.aes = list(size = 4)
    )
  ) +

  scale_linetype_manual(
    values = c(
      "Statistically significant\n(p < 0.05)" = "solid",
      "Kano State boundary" = "solid"
    ),

    breaks = c(
      "Statistically significant\n(p < 0.05)",
      "Kano State boundary"
    ),

    name = NULL,

    guide = guide_legend(order = 3)
  ) +

  ggspatial::annotation_scale(
    location = "bl",
    width_hint = 0.22,
    text_cex = 0.7,
    line_width = 0.6
  ) +

  ggspatial::annotation_north_arrow(
    location = "br",
    which_north = "true",

    pad_x = unit(0.18, "in"),
    pad_y = unit(0.18, "in"),

    style = north_arrow_fancy_orienteering(
      text_size = 6
    )
  ) +

  labs(
    title = "Maternal Death Risk Surface — Kano State, Nigeria",

    subtitle =
      "Kernel density ratio (adaptive bandwidth) | Contours: p < 0.05"
  ) +

  coord_sf(expand = FALSE) +

  theme_void(base_size = 12) +

  theme(

    plot.title = element_text(
      family = "serif",
      face = "bold",
      size = 20,
      hjust = 0.5,
      margin = margin(b = 4)
    ),

    plot.subtitle = element_text(
      family = "serif",
      size = 13,
      color = "grey30",
      hjust = 0.5,
      margin = margin(b = 8)
    ),

    legend.position = "right",
    legend.box = "vertical",
    legend.box.just = "left",
    legend.spacing.y = unit(0.45, "cm"),
    legend.key.height = unit(0.9, "cm"),
    legend.key.width = unit(1.3, "cm"),
    legend.margin = margin(0, 8, 0, 8),

    legend.title = element_text(
      family = "serif",
      face = "bold",
      size = 10
    ),

    legend.text = element_text(
      family = "serif",
      size = 10
    ),

    panel.background = element_rect(
      fill = "white",
      color = NA
    ),

    plot.background = element_rect(
      fill = "white",
      color = NA
    ),

    plot.margin = margin(
      t = 14,
      r = 18,
      b = 12,
      l = 18
    )
  )

print(p_risk)


dir.create("outputs", showWarnings = FALSE)

ggsave(
  filename = "outputs/maternal_risk_surface_kano_publication.png",
  plot = p_risk,
  width = 11,
  height = 7.5,
  dpi = 600,
  bg = "white"
)
```




## Step 8 — Normalize risk surface

```{r normalize, include=FALSE}
# Min-max normalization to [0, 1]
rr_min  <- terra::global(rr_raster, "min", na.rm = TRUE)$min
rr_max  <- terra::global(rr_raster, "max", na.rm = TRUE)$max
rr_norm <- (rr_raster - rr_min) / (rr_max - rr_min)

cat("Normalized range: [",
    terra::global(rr_norm, "min", na.rm = TRUE)$min, ",",
    terra::global(rr_norm, "max", na.rm = TRUE)$max, "]\n")
```

## Step 9 — Extract risk at model grid centroids

Risk values are extracted at each model grid cell centroid.
This aligns the coarse risk raster (~500m) with the finer model grid,
avoiding resolution mismatch problems that arise from rasterizing in the
opposite direction.

```{r extract-risk-at-model, include=FALSE}
# Compute centroids of model grid polygons (Mollweide)
model_pts     <- sf::st_centroid(model_proj)

# Extract normalized risk value at each centroid
risk_at_model <- terra::extract(rr_norm, terra::vect(model_pts))[, 2]

cat("Model grid cells:", length(risk_at_model), "\n")
cat("Cells with valid risk:", sum(!is.na(risk_at_model)), "\n")
```


# Comparative Analysis

## Risk surface vs Accessibility Score — density comparison

Each variable is extracted independently: risk from the raster, accessibility from
the model grid attribute table. Log-transformation corrects right skew in accessibility.

```{r density-comparison}
# Risk: valid pixels from normalized raster
rr_vec <- as.vector(terra::values(rr_norm, na.rm = TRUE))

# Accessibility: directly from model attribute table, log-normalised
acc_valid <- model_proj$Accessibility_Index_Standard
acc_valid <- acc_valid[!is.na(acc_valid)]
acc_log   <- log(acc_valid + 1e-10)
acc_vec   <- (acc_log - min(acc_log)) / (max(acc_log) - min(acc_log))

df_combined <- data.frame(
  value  = c(rr_vec, acc_vec),
  source = rep(c("Risk Surface", "Accessibility Score"),
               c(length(rr_vec), length(acc_vec)))
)

neutral_val <- (0 - rr_min) / (rr_max - rr_min)

ggplot(df_combined, aes(x = value, fill = source)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = neutral_val, linetype = "dashed", color = "black") +
  annotate("text", x = neutral_val, y = Inf, vjust = 2, hjust = -0.1,
           label = "log RR = 0\n(no excess risk)", size = 3.2, color = "grey30") +
  scale_fill_manual(values = c("Risk Surface"        = "orange",
                               "Accessibility Score" = "steelblue")) +
  labs(
    title    = "Risk Surface vs Accessibility Score — Kano, Nigeria",
    subtitle = "Risk = normalized log RR | Accessibility log-transformed | Dashed line = no excess risk",
    x = "Normalized Value", y = "Density", fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```


## Scatter plot — grid cell level correlation

```{r scatter-correlation}
vals <- data.frame(
  risk = risk_at_model,
  acc  = model_proj$Accessibility_Index_Standard
) |> na.omit()

r_val  <- round(cor(vals$risk, vals$acc, method = "spearman"), 3)
r2_val <- round(r_val^2, 3)

ggplot(vals, aes(x = risk, y = acc)) +
  geom_point(alpha = 0.3, size = 1.2, color = "#2E86AB") +
  geom_smooth(method = "lm", color = "#E84855", linewidth = 1, se = TRUE) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -1,
           label = paste0("ρ = ", r_val, "\nρ² = ", r2_val),
           size = 4.5, color = "grey20") +
  labs(
    title    = "Mortality Risk vs Accessibility — grid cell level",
    subtitle = "Each point = one model grid cell centroid | Spearman correlation",
    x = "Normalized Log Risk Ratio", y = "Accessibility Score"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40"))
```


## Risk by deprivation category

```{r risk-by-deprivation}
df_dep <- data.frame(
  risk        = risk_at_model,
  deprivation = model_proj$result
) |>
  na.omit() |>
  dplyr::mutate(deprivation = factor(deprivation,
    levels = 0:2,
    labels = c("Low Deprivation", "Medium Deprivation", "High Deprivation")
  ))

cat("Cell counts:\n"); print(table(df_dep$deprivation))

kw <- kruskal.test(risk ~ deprivation, data = df_dep)
cat("\nKruskal-Wallis p =", kw$p.value, "\n")
cat("\nMedian risk by group:\n")
print(tapply(df_dep$risk, df_dep$deprivation, median))

ggplot(df_dep, aes(x = deprivation, y = risk, fill = deprivation)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_fill_manual(values = c(
    "Low Deprivation"    = "steelblue",
    "Medium Deprivation" = "orange",
    "High Deprivation"   = "#D6604D"
  )) +
  annotate("text", x = 2, y = 0.95,
           label = paste0("Kruskal-Wallis\np = ", format(kw$p.value, digits = 3)),
           size = 3.5) +
  labs(
    title = "Maternal Death Risk by Deprivation Category — Kano, Nigeria",
    x = "", y = "Normalized Risk Surface", fill = ""
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title      = element_text(face = "bold", size = 12),
        legend.position = "none")
```

### Pairwise Dunn test (Bonferroni correction)

```{r dunn-test}
dunn.test(df_dep$risk, df_dep$deprivation, method = "bonferroni")
```

---
# Spatial co-clustering

## Prepare shared LISA data
```{r prepare-lisa-data}
# Convert risk raster pixels to polygons — these are the analysis units
# for spatial autocorrelation. Resolution ~500m, matching the effective
# resolution of the kernel density estimate.
lisa_poly <- terra::as.polygons(rr_raster, dissolve = FALSE)
names(lisa_poly)[1] <- "risk"

# Aggregate the finer model grid's accessibility values onto each risk
# pixel (mean of all model cells overlapping a pixel). This puts both
# variables on the same spatial scale — essential, otherwise the LISA
# would falsely inflate autocorrelation by inserting identical risk
# values into many small grid cells.
acc_rast <- terra::rasterize(
  terra::vect(model_proj), rr_norm,
  field = "Accessibility_Index_Standard"
)
lisa_poly$acc <- terra::extract(acc_rast, lisa_poly,
                                fun = mean, na.rm = TRUE)[, 2]

# Drop pixels with NA in either variable (typically edge pixels)
lisa_sf <- sf::st_as_sf(lisa_poly) |> na.omit()

# Invert accessibility: now HIGH value = HIGH deprivation, matching
# the direction of "risk". Without this, the bivariate Moran's I would
# come out NEGATIVE for what is intuitively a positive co-cluster.
lisa_sf$acc_inv <- 1 - lisa_sf$acc

cat("Pixels retained for spatial analysis:", nrow(lisa_sf), "\n")
```

## Global Moran's I — univariate and bivariate
Three Moran's I statistics are reported:

1. **Univariate Moran's I for risk** — is mortality risk itself
   spatially clustered? (Expected: yes, risk is rarely random in space)
2. **Univariate Moran's I for deprivation** — is the model output
   spatially clustered? (Expected: yes, accessibility has obvious
   spatial structure)
3. **Bivariate Moran's I (risk × deprivation)** — do they cluster
   *together*? This is the global counterpart of the LISA map and is
   the headline number for the spatial validation.
   
```{r global-morans-i}
# ---- Build neighbor list (queen contiguity) ----
# Queen contiguity: two pixels are neighbors if they share an edge OR
# a corner. Same definition used in the bivariate LISA below — they
# must match for the two analyses to be comparable.
nb <- spdep::poly2nb(lisa_sf, queen = TRUE)
lw <- spdep::nb2listw(nb, style = "W", zero.policy = TRUE)
# style = "W" → row-standardized weights (each row sums to 1)
# zero.policy = TRUE → tolerate isolated pixels (edge effects)

cat("Average neighbours per pixel:",
    round(mean(spdep::card(nb)), 2), "\n\n")

# ---- (a) Univariate Moran's I for risk ----
moran_risk <- spdep::moran.test(lisa_sf$risk, lw,
                                 zero.policy = TRUE,
                                 randomisation = TRUE)

# ---- (b) Univariate Moran's I for deprivation (inverted accessibility) ----
moran_acc <- spdep::moran.test(lisa_sf$acc_inv, lw,
                                zero.policy = TRUE,
                                randomisation = TRUE)

# ---- (c) Bivariate Global Moran's I ----
# Definition: I_xy = Σ(x_i * Σ(w_ij * y_j)) / Σ(x_i²)
# i.e. correlation between standardized x and spatial-lag of standardized y.
x  <- scale(lisa_sf$risk)[, 1]
y  <- scale(lisa_sf$acc_inv)[, 1]
Wy <- spdep::lag.listw(lw, y, zero.policy = TRUE)
biv_I <- sum(x * Wy) / sum(x^2)

# Monte Carlo permutation test:
# Randomly shuffle y 999 times, recompute the statistic each time,
# and see how often a shuffled value reaches the observed magnitude.
# This is the standard inference approach for bivariate Moran's I
# (closed-form distributions don't exist for the bivariate case).
set.seed(42)
n_perm <- 999
biv_perm <- replicate(n_perm, {
  y_shuf  <- sample(y)
  Wy_shuf <- spdep::lag.listw(lw, y_shuf, zero.policy = TRUE)
  sum(x * Wy_shuf) / sum(x^2)
})
biv_p <- (sum(abs(biv_perm) >= abs(biv_I)) + 1) / (n_perm + 1)

# ---- Paper-ready output ----
cat("=========================================================\n")
cat("GLOBAL MORAN'S I RESULTS\n")
cat("=========================================================\n\n")

cat("Univariate Moran's I — Mortality risk\n")
cat(sprintf("  I = %.4f   E[I] = %.4f   p = %.3g\n\n",
            moran_risk$estimate["Moran I statistic"],
            moran_risk$estimate["Expectation"],
            moran_risk$p.value))

cat("Univariate Moran's I — Deprivation (inverted accessibility)\n")
cat(sprintf("  I = %.4f   E[I] = %.4f   p = %.3g\n\n",
            moran_acc$estimate["Moran I statistic"],
            moran_acc$estimate["Expectation"],
            moran_acc$p.value))

cat("Bivariate Moran's I — Risk × Deprivation\n")
cat(sprintf("  I = %.4f   p = %.3g   (999 permutations)\n",
            biv_I, biv_p))
cat("=========================================================\n")
```

## Bivariate Local Moran's I — Risk × Accessibility

Analysis units are the risk raster pixels (~500m), which match the effective
spatial resolution of the kernel density estimate. Accessibility is aggregated
(mean) from the finer model grid onto each risk pixel. This ensures both
variables share the same spatial scale, avoiding artificial autocorrelation
from inserting identical risk values into many smaller grid cells.

```{r bivariate-lisa-map}
# Queen contiguity weights for rgeoda (same definition as spdep above)
queen_w <- rgeoda::queen_weights(lisa_sf)

# Bivariate Local Moran's I: risk (x) vs inverted accessibility (y)
biv_lisa <- rgeoda::local_bimoran(queen_w, lisa_sf[, c("risk", "acc_inv")])

# Attach cluster labels to the spatial object
lisa_sf$cluster <- factor(
  rgeoda::lisa_clusters(biv_lisa, cutoff = 0.05),
  levels = c(1, 2, 3, 4, 0),
  labels = c("High-High", "Low-Low", "Low-High", "High-Low", "Not Significant")
)

cat("\nCluster counts:\n"); print(table(lisa_sf$cluster))

# Reproject to WGS84 for plotting (ggplot/sf works best in WGS84 for maps)
lisa_wgs84 <- sf::st_transform(lisa_sf, crs = 4326)

cluster_colors <- c(
  "High-High"       = "#D6604D",
  "Low-Low"         = "#4393C3",
  "Low-High"        = "#92C5DE",
  "High-Low"        = "#F4A582",
  "Not Significant" = "grey85"
)

ggplot() +
  geom_sf(data = lisa_wgs84, aes(fill = cluster), color = NA) +
  geom_sf(data = sf::st_transform(kano, crs = 4326),
          fill = NA, color = "black", linewidth = 0.5) +
  scale_fill_manual(values = cluster_colors, name = "Bivariate\nLISA") +
  ggspatial::annotation_scale(location = "bl", width_hint = 0.2) +
  ggspatial::annotation_north_arrow(
    location = "br",
    style    = north_arrow_fancy_orienteering(text_size = 6)
  ) +
  guides(fill = guide_legend(override.aes = list(color = NA))) +
  labs(
    title    = "Bivariate LISA — Maternal Death Risk vs Accessibility Deprivation",
    subtitle = "High-High = high risk + high deprivation (priority areas) | p < 0.05"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey40", size = 9),
    legend.key.size = unit(0.8, "cm")
  )

ggsave(
  filename = "outputs/bivariate_lisa_map.png",
  width = 11,
  height = 8,
  dpi = 600,
  bg = "white"
)
```

# Save Outputs

```{r save-outputs, include=FALSE}
dir.create("outputs", showWarnings = FALSE)

# Reproject risk raster to WGS84 and resample to model grid resolution
rr_wgs84 <- terra::project(rr_raster, "EPSG:4326")

res_x    <- mean(model_outputs$lon_max - model_outputs$lon_min)
res_y    <- mean(model_outputs$lat_max - model_outputs$lat_min)
template <- terra::rast(
  xmin       = min(model_outputs$lon_min),
  xmax       = max(model_outputs$lon_max),
  ymin       = min(model_outputs$lat_min),
  ymax       = max(model_outputs$lat_max),
  resolution = c(res_x, res_y),
  crs        = "EPSG:4326"
)

rr_resampled <- terra::resample(rr_wgs84, template, method = "bilinear")
cat("Saved raster resolution:", terra::res(rr_resampled), "\n")

# PNG map
png("outputs/risk-surface-kano.png", width = 1200, height = 900, res = 150)
print(p_risk)
dev.off()

# GeoTIFF (WGS84, model grid resolution)
terra::writeRaster(rr_resampled, "outputs/risk-surface-kano.tif", overwrite = TRUE)
```




