---
title: "Estimating Risk Surface"
output: html_notebook
---

# Preliminary tasks 

## Preparation
The code processes the healthcare model outputs to estimate a 
risk surface based on https://arxiv.org/pdf/1707.06888
Read CSV file

### Required packages
To secure the necessary packages, run the following code:
```{r include=FALSE}
install.packages('sparr')
```
### Load required packages
```{r include=FALSE}
library(sf)
library(sparr)
library(jsonlite)
library(sp)
library(dplyr)
library(ggplot2)
```

## View first few rows
head(data)
```{r read-model-ouputs, include=FALSE}
# Read CSV file
data <- read.csv("data/model-output-acc-score.csv")
head(data)
```

## View first few rows
```{r read-maternal-deaths, include=FALSE}
maternal_deaths <- sf::st_read("data/maternaldeaths.geojson", quiet = TRUE)
head(maternal_deaths)
```

## View first few rows
```{r read-study-area, include=FALSE}
# Read CSV file
kano <- sf::st_read("data/grid-boundary-kano.gpkg", quiet = TRUE)
head(kano)
```

# Default estimation using example

Estimate risk surface using the PBC dataset - Available via the sparr package

This section needs to be reviewed. With the  
```{r estimate-initial-pbc, include=FALSE}
data(pbc)
pbccas <- split(pbc)$case
pbccon <- split(pbc)$control
```

```{r estimate-initial-pbc, include=FALSE}
risk(f=pbccas,g=pbccon,doplot=TRUE)
```

```{r estimate-pbc-geoemtric, include=FALSE}
h0 <- OS(pbc, nstar="geometric")
pbc_rr <- risk(f=pbccas, g=pbccon, h0=h0, adapt=TRUE, tolerate=TRUE, hp=OS(pbc)/2, pilot.symmetry="pooled", davies.baddeley=0.05)
plot(pbc_rr)
```


## Estimate risk surface using the maternal deaths dataset


We need to create the two datasets, control dataset will come from the population
dataset. It will use a function created to generate random points within the study area.

### Population at risk
The paper refers to population at risk data. Here we will use WorldPop data for women
of childbearing age in Kano, Nigeria.
Note that the study area is constrained to the area defined by the IDEAMAPS
data ecosystem project. 
```{r generate-random-population-points, include=FALSE}
source("createrandompoints.R")
raster_to_random_points("data/Kano_Women_15-49_WorldPOP.tif", "outputs/random-points-women-kano.gpkg")
# Read the geospatial data from random points file
womenrepage <- sf::st_read("outputs/random-points-women-kano.geopkg", quiet = TRUE)


```

Build the spatial window for the PPP objects
```{r read-random-population-points, include=FALSE}
maternal_deaths <- sf::st_read("outputs/random-points-women-kano.gpkg", quiet = TRUE)
# Create spatial window (observation region)
# Get the boundary from the kano sf object
# kano_boundary <- st_union(kano)  # Combine all geometries into one
# kano_boundary_sp <- as(kano_boundary, "Spatial")  # Convert to Spatial object
# kano_boundary_owin <- as.owin(kano_boundary_sp)  # Convert to owin object

# Create spatial window (observation region)
# Get bounding box and create an owin object
bbox <- st_bbox(maternal_deaths)
window <- owin(xrange = c(bbox["xmin"], bbox["xmax"]), 
               yrange = c(bbox["ymin"], bbox["ymax"]))

```

Once the data is generated from the raster dataset the variable is set
```{r read-random-population-points, include=FALSE}

# from geojson to ppp
deaths <- st_as_sf(maternal_deaths)
# Extract coordinates from sf object
coords <- st_coordinates(deaths)

# Create ppp object
deaths_ppp <- ppp(x = coords[,1], y = coords[,2], window = window)

# View summary
summary(deaths_ppp)

```


### Cases
The Cases
```{r read-mortality-cases, include=FALSE}


population <- st_as_sf(womenrepage)
# Extract coordinates from sf object
coords <- st_coordinates(population)

# Create ppp object
pop_ppp <- ppp(x = coords[,1], y = coords[,2], window = window)

# View summary
summary(pop_ppp)
```


### Estimating risk surface using the PBC dataset - Available via the sparr package
```{r estimate-initial-pbc, include=FALSE}
hfix <- LSVC.risk(f = deaths_ppp, g = pop_ppp)
#risk(f = deaths_ppp, g = pop_ppp, window = window, doplot = TRUE)
```

```{r estimate-pbc-geoemtric, include=FALSE}
h0 <- OS(pbc, nstar="geometric")
pbc_rr <- risk(f=pbccas, g=pbccon, h0=h0, adapt=TRUE, tolerate=TRUE, hp=OS(pbc)/2, pilot.symmetry="pooled", davies.baddeley=0.05)
plot(pbc_rr)
```


