---
title: "Estimating Risk Surface"
output:
  html_document:
    df_print: paged
---

# Preliminary tasks 

## Preparation
This notebook estimates a spatial risk surface for maternal deaths in Kano, Nigeria,
based on the kernel density ratio method: https://arxiv.org/pdf/1707.06888

Key optimizations applied:
- WorldPop raster is aggregated before point generation to reduce memory load
- Population points are capped at a maximum sample size (10,000)
- Fixed bandwidth used first for speed; adaptive bandwidth optional
- Duplicate chunk names fixed
- Cases/Controls variable naming corrected
- All data reprojected to Mollweide (ESRI:54009) for equal-area analysis

### Install required packages
```{r install-packages, include=FALSE, eval=FALSE}
# Only run once - set eval=FALSE after first run
install.packages(c('sparr', 'terra', 'sf', 'spatstat', 'dplyr', 'ggplot2', "tidyterra", "ggspatial", "viridis", "cowplot", "rlang"))
```

### Load required packages
```{r load-packages, include=FALSE}
library(sf)
library(sparr)
library(terra)       # Modern replacement for raster package
library(spatstat)
library(dplyr)
library(ggplot2)
library(tidyterra)
library(ggspatial)
library(cowplot)
library(viridis)
```
---

# Load Data

## Model outputs
```{r read-model-outputs, include=FALSE}
data <- read.csv("data/model-output-acc-score.csv")
head(data)
```

## Maternal deaths (Cases)
```{r read-maternal-deaths, include=FALSE}
maternal_deaths <- sf::st_read("data/maternaldeaths.geojson", quiet = TRUE)
head(maternal_deaths)
```

## Study area boundary
```{r read-study-area, include=FALSE}
kano <- sf::st_read("data/grid-boundary-kano.gpkg", quiet = TRUE)
head(kano)
```

## Population data (Women of childbearing age 15-49, 100m resolution)
```{r read-population, include=FALSE}
population <- sf::st_read("data/pop-grid-kano-centroids.gpkg", quiet = TRUE)
head(population)
```

---

# Main Analysis: Maternal Death Risk Surface in Kano

## Step 1: Reproject all data to Mollweide
#
# Mollweide (ESRI:54009) is an equal-area projection suitable for spatial analysis.
# spatstat's as.owin() requires a projected coordinate system (metres), not WGS84 degrees.
# All datasets must use the same CRS for ppp objects to align correctly.

```{r reproject-data}
CRS_MOLLWEIDE <- "ESRI:54009"

kano_proj            <- sf::st_transform(kano,            crs = CRS_MOLLWEIDE)
maternal_deaths_proj <- sf::st_transform(maternal_deaths, crs = CRS_MOLLWEIDE)
population_proj      <- sf::st_transform(population,      crs = CRS_MOLLWEIDE)
```

## Step 2: Filter maternal deaths to Kano study area
#
# maternal_deaths contains records from the wider Kano region.
# Deaths outside the study area boundary have no corresponding
# population Controls and would bias the risk surface upward.

```{r filter-deaths}
maternal_deaths_kano <- sf::st_filter(maternal_deaths_proj, kano_proj)
cat("Deaths within Kano study area:", nrow(maternal_deaths_kano),
    "out of", nrow(maternal_deaths_proj), "\n")

# Visual check
plot(sf::st_geometry(kano_proj), main = "Maternal Deaths within Kano Study Area")
plot(sf::st_geometry(maternal_deaths_kano), add = TRUE,
     col = "red", pch = 16, cex = 1.2)
```

## Step 3: Generate population points from 100m grid polygons
#
# Each polygon represents a 100m grid cell with a population value.
# Random points are generated within each polygon proportional to population.
# SCALE_FACTOR reduces total point count while preserving relative spatial distribution.
# st_sample() places points randomly inside each polygon - no manual cell size needed.


```{r generate-population-points}
# Filter out NA and zero population cells
pop_valid <- population_proj[!is.na(population_proj$pop) & population_proj$pop > 0, ]
cat("Cells with population > 0:", nrow(pop_valid), "\n")
cat("Population summary:\n")
summary(pop_valid$pop)

# Scale down population values proportionally
SCALE_FACTOR <- 10
pop_valid[["n_points"]] <- round(pop_valid[["pop"]] / SCALE_FACTOR)

# Filter out cells that round down to 0
pop_valid <- pop_valid[!is.na(pop_valid[["n_points"]]) & pop_valid[["n_points"]] > 0, ]
cat("Cells to process:", nrow(pop_valid), "\n")
cat("Total points to generate:", sum(pop_valid[["n_points"]]), "\n")

# Generate random points within each polygon
set.seed(42)
points_list <- lapply(1:nrow(pop_valid), function(i) {
  n    <- pop_valid[["n_points"]][i]
  bbox <- sf::st_bbox(pop_valid[i, ])
  data.frame(
    x = runif(n, bbox["xmin"], bbox["xmax"]),
    y = runif(n, bbox["ymin"], bbox["ymax"])
  )
})

points_df   <- do.call(rbind, points_list)
womenrepage <- sf::st_as_sf(points_df, coords = c("x", "y"),
                             crs = sf::st_crs(population_proj))

cat("Population points generated:", nrow(womenrepage), "\n")
```



## Step 4: Build spatial window from Kano polygon boundary
#
# Using the actual Kano boundary polygon (not a bounding box rectangle)
# ensures the risk surface matches the study area shape exactly.
# This requires projected coordinates - hence the UTM reprojection above.


```{r build-window}
kano_poly   <- sf::st_cast(kano_proj, "POLYGON")
kano_window <- spatstat.geom::as.owin(kano_poly)
```
## Step 5: Create PPP objects

```{r create-cases-ppp}
# Cases: maternal deaths within Kano study area
deaths_coords <- sf::st_coordinates(maternal_deaths_kano)
deaths_ppp    <- spatstat.geom::ppp(
  x      = deaths_coords[, 1],
  y      = deaths_coords[, 2],
  window = kano_window
)
```

```{r create-controls-ppp}
# Controls: population at risk (women of childbearing age)
pop_coords <- sf::st_coordinates(womenrepage)
pop_ppp    <- spatstat.geom::ppp(
  x      = pop_coords[, 1],
  y      = pop_coords[, 2],
  window = kano_window
)
```

## Step 6: Estimate risk surface

### Option A: Fixed bandwidth (run first to verify everything works)
```{r risk-fixed}
rr_fixed <- risk(f = deaths_ppp, g = pop_ppp, doplot = TRUE)
plot(rr_fixed, main = "Risk Surface - Fixed Bandwidth")
```

### Option B: Geometric bandwidth (recommended)
```{r risk-geometric}
h0 <- OS(deaths_ppp, nstar = "geometric")
cat("Optimal bandwidth:", h0, "\n")

rr_geom <- risk(
  f        = deaths_ppp,
  g        = pop_ppp,
  h0       = h0,
  tolerate = TRUE,
  doplot   = TRUE
)
plot(rr_geom, main = "Risk Surface - Geometric Bandwidth")

```

### Option C: Adaptive bandwidth (most accurate, slowest)
# Set eval=TRUE in chunk header when ready to run - may take 10-30 mins
```{r risk-adaptive, eval=FALSE}
hpilot.f <- BOOT.density(deaths_ppp)
hpilot.g <- BOOT.density(pop_ppp)

rr_adapt <- risk(
  f               = deaths_ppp,
  g               = pop_ppp,
  h0              = h0,
  adapt           = TRUE,
  tolerate        = TRUE,
  hp              = c(hpilot.f, hpilot.g),
  pilot.symmetry  = "pooled",
  davies.baddeley = 0.05
)
plot(rr_adapt, main = "Risk Surface - Adaptive Bandwidth")
```


```{r}
# Prepare data for ggplot

# Convert rr_adapt risk surface to terra raster
rr_raster <- terra::rast(rr_adapt$rr)
terra::crs(rr_raster) <- "ESRI:54009"

# Convert significance surface (p-values)
rr_p <- terra::rast(rr_adapt$P)
terra::crs(rr_p) <- "ESRI:54009"

# Create significance mask (p < 0.05)
sig_mask <- rr_p < 0.05

# Convert kano boundary to sf for plotting
kano_sf <- sf::st_as_sf(kano_proj)

# ============================================================
# Main risk surface plot
p_risk <- ggplot() +
  tidyterra::geom_spatraster(data = rr_raster) +
  scale_fill_gradientn(
    colours  = c("#000080", "#0000FF", "#00FFFF", "#FFFF00", "#FF8000", "#FF0000"),
    na.value = "transparent",
    name     = "Log Risk\nRatio"
  ) +
  geom_sf(data = kano_sf,
          fill      = NA,
          color     = "black",
          linewidth = 0.5) +
  tidyterra::geom_spatraster_contour(
    data      = rr_p,
    breaks    = 0.05,
    color     = "black",
    linewidth = 0.6
  ) +
  geom_sf(data   = maternal_deaths_kano,
          color  = "black",
          fill   = "yellow",
          shape  = 21,
          size   = 1.5,
          stroke = 0.4) +
  ggspatial::annotation_scale(
    location   = "bl",
    width_hint = 0.2,
    text_cex   = 0.7
  ) +
  ggspatial::annotation_north_arrow(
    location    = "br",
    which_north = "true",
    pad_x       = unit(0.1, "in"),
    pad_y       = unit(0.1, "in"),
    style       = north_arrow_fancy_orienteering(text_size = 6)
  ) +
  labs(
    title    = "Maternal Death Risk Surface — Kano, Nigeria",
    subtitle = "Kernel density ratio (adaptive bandwidth)\nBlack contours indicate p < 0.05 significance"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle   = element_text(size = 9, color = "grey40", hjust = 0.5),
    legend.position = "right",
    legend.title    = element_text(size = 8, face = "bold"),
    legend.text     = element_text(size = 7),
    plot.margin     = margin(10, 20, 10, 20)
  )

print(p_risk)
```
```{r}

```


## Step 7: Resample risk surface to match model output resolution
#
# rr_adapt is computed in Mollweide (ESRI:54009).
# Convert to WGS84 for compatibility with other datasets and GIS software.
# No resampling - bandwidth (~4747m) determines the natural resolution of
# the risk surface regardless of input data resolution.
```{r save-risk-surface}
# Convert to terra raster and set Mollweide CRS
rr_raster <- terra::rast(rr_adapt$rr)
terra::crs(rr_raster) <- "ESRI:54009"

# Also extract significance surface
rr_p <- terra::rast(rr_adapt$P)
terra::crs(rr_p) <- "ESRI:54009"

# Reproject to WGS84
rr_wgs84 <- terra::project(rr_raster, "EPSG:4326")

# Resample risk surface to 100m and save GeoTIFF
# Extract resolution from model output grid
res_x <- mean(data$lon_max - data$lon_min)
res_y <- mean(data$lat_max - data$lat_min)

# Create template matching model output grid exactly
template <- terra::rast(
  xmin       = min(data$lon_min),
  xmax       = max(data$lon_max),
  ymin       = min(data$lat_min),
  ymax       = max(data$lat_max),
  resolution = c(res_x, res_y),
  crs        = "EPSG:4326"
)

# Resample to 100m
rr_resampled <- terra::resample(rr_wgs84, template, method = "bilinear")
```

## Step 8: Save as GeoTIFF and PNG
```{r}
# Save sparr plot as PNG
png("outputs/risk-surface-kano.png", width = 1200, height = 900, res = 150)
plot(p_risk, main = "Maternal Death Risk Surface - Kano, Nigeria")
dev.off()

# Save raw raster as GeoTIFF
terra::writeRaster(rr_resampled,
                   "outputs/risk-surface-kano.tiff",
                   overwrite = TRUE)
```


## make same resolution, 1km and 100m
check numeric of accessibility score
check the categories of high medium and low (thresholds)
maybe also try general healthcare model

Further analysis :
1. Bi-variate cluster analysis 
2. Test only with the built up area

### calculate avg risk of high, medium and low
### compare the number, regression
### statistic analysis, max, min... volin plot
### Bi-variate cluster analysis, check different between clusters
### analysis build up area


## Step 9: Normalize risk surface to 0-1

```{r normalize}
rr_min  <- terra::global(rr_resampled, "min", na.rm = TRUE)$min
rr_max  <- terra::global(rr_resampled, "max", na.rm = TRUE)$max
rr_norm <- (rr_resampled - rr_min) / (rr_max - rr_min)

cat("Normalized min:", terra::global(rr_norm, "min", na.rm = TRUE)$min, "\n")
cat("Normalized max:", terra::global(rr_norm, "max", na.rm = TRUE)$max, "\n")
```

## Step 10: Density plot comparison - Risk Surface vs Accessibility Score

```{r density-comparison}
# Extract normalized risk surface values
rr_vec <- as.vector(terra::values(rr_norm, na.rm = TRUE))

# Log-transform Accessibility_standard to address extreme right skew
# (MinMaxScaler preserves skew - log transform improves visualization)
acc_log  <- log(data$Accessibility_standard + 1e-10)
acc_min  <- min(acc_log, na.rm = TRUE)
acc_max  <- max(acc_log, na.rm = TRUE)
acc_vec  <- as.vector((acc_log - acc_min) / (acc_max - acc_min))

# Combined density plot
df_combined <- data.frame(
  value  = c(rr_vec, acc_vec),
  source = c(rep("Risk Surface",        length(rr_vec)),
             rep("Accessibility Score", length(acc_vec)))
)

p_compare <- ggplot(df_combined, aes(x = value, fill = source)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("Risk Surface"        = "orange",
                               "Accessibility Score" = "steelblue")) +
  labs(
    title    = "Risk Surface vs Accessibility Score - Kano, Nigeria",
    subtitle = "Both normalized to 0-1 (Accessibility Score log-transformed for visualization)",
    x        = "Normalized Value (0-1)",
    y        = "Density",
    fill     = ""
  ) +
  theme_minimal() +
  theme(legend.position = "top")

print(p_compare)
cat("Density plot saved\n")
```
do the correlation analysis
calculate the number of maternity death in different categories
new model vision (improve the thresholds)- check with par result and maternal death
compare two vision of model outputs



First, absolute deaths / categories
Second, Risk surface comparison
Third,
New model version? new thresholds and parameters?
New comparison with the validation data

scatter plot, x will be accessbility score, y is risk surface


