---
title: "Estimating Risk Surface"
output:
  html_document:
    df_print: paged
---

# Preliminary tasks 

## Preparation
This notebook estimates a spatial risk surface for maternal deaths in Kano, Nigeria,
based on the kernel density ratio method: https://arxiv.org/pdf/1707.06888

Key optimisations applied:
- WorldPop raster is aggregated before point generation to reduce memory load
- Population points are capped at a maximum sample size (10,000)
- Fixed bandwidth used first for speed; adaptive bandwidth optional
- Duplicate chunk names fixed
- Cases/Controls variable naming corrected
- All data reprojected to Mollweide (ESRI:54009) for equal-area analysis

### Install required packages

There is a selection of packages required for the risk surface analysys. Among them, geospatial data management packages and "sparr".
```{r install-packages, include=FALSE, eval=FALSE}
# Run once, then set eval=FALSE
install.packages(c(
  "sparr", "terra", "sf", "spatstat", "dplyr", "ggplot2",
  "tidyterra", "ggspatial", "viridis", "cowplot", "rlang",
  "spdep", "rgeoda", "dunn.test"
))
```

### Load required packages
```{r load-packages, include=FALSE}
library(sf)
library(sparr)
library(terra)       # Modern replacement for raster package
library(spatstat)
library(dplyr)
library(ggplot2)
library(tidyterra)
library(ggspatial)
library(cowplot)
library(viridis)
library(spdep)     # Added to estimate local Moran's and bivariate local Moran's
library(rgeoda)
library(dunn.test)
```
---

# Load Data

## Model outputs
This data set contains the outputs for the "Emergency Obstetric Care Access Deprivation" model generated by the IDEAMAPS data ecosystem.
```{r read-model-outputs, include=FALSE}
data <- read.csv("data/model-output-classification.csv")
head(data)
```

## Model outputs
load geopackage
```{r model-outputs, include=FALSE}
model_outputs <- sf::st_read("data/model-output-classification.gpkg", quiet = TRUE)
head(model_outputs)
```

## Maternal deaths (Cases)
This data, provided by the KASSEP project, compiles cases of maternal deaths in Kano
```{r read-maternal-deaths, include=FALSE}
maternal_deaths <- sf::st_read("data/maternaldeaths.geojson", quiet = TRUE)
head(maternal_deaths)
```

## Study area boundary
```{r read-study-area, include=FALSE}
kano <- sf::st_read("data/grid-boundary-kano.gpkg", quiet = TRUE)
head(kano)
```

## Population data (Women of childbearing age 15-49, 100m resolution)
```{r read-population, include=FALSE}
population <- sf::st_read("data/pop-grid-kano-centroids.gpkg", quiet = TRUE)
head(population)
```

---

# Main Analysis

## Step 1 — Reproject to Mollweide

Equal-area projection required by `spatstat`; all layers must share the same CRS.

```{r reproject, include=FALSE}
CRS_MOLLWEIDE <- "ESRI:54009"

kano_proj            <- sf::st_transform(kano,            crs = CRS_MOLLWEIDE)
maternal_deaths_proj <- sf::st_transform(maternal_deaths, crs = CRS_MOLLWEIDE)
population_proj      <- sf::st_transform(population,      crs = CRS_MOLLWEIDE)
model_proj           <- sf::st_transform(model_outputs,   crs = CRS_MOLLWEIDE)
```

## Step 2 — Clip deaths to 10 km study-area buffer

Deaths outside the study boundary have no matching population controls
and would inflate the risk surface.

```{r clip-deaths}
kano_buffer         <- sf::st_buffer(kano_proj, dist = 10000)
maternal_deaths_kano <- sf::st_filter(maternal_deaths_proj, kano_buffer)

cat("Deaths retained:", nrow(maternal_deaths_kano),
    "of", nrow(maternal_deaths_proj), "\n")

# Quick visual check
plot(sf::st_geometry(kano_buffer),
     main = "Maternal Deaths — 10 km Buffer", border = "grey50", lty = 2)
plot(sf::st_geometry(kano_proj),        add = TRUE, border = "black")
plot(sf::st_geometry(maternal_deaths_kano), add = TRUE, col = "red", pch = 16, cex = 0.8)
```


## Step 3: Generate population points from 100m grid polygons

Each polygon represents a 100m grid cell with a population value.
Random points are generated within each polygon proportional to population.
SCALE_FACTOR reduces total point count while preserving relative spatial distribution.
st_sample() places points randomly inside each polygon - no manual cell size needed.

Points are placed randomly inside each 100 m grid cell, proportional to population
(`SCALE_FACTOR = 10` reduces total count while preserving spatial distribution).

```{r generate-pop-points}
pop_valid <- population_proj |>
  dplyr::filter(!is.na(pop), pop > 0) |>
  dplyr::mutate(n_points = round(pop / 10)) |>
  dplyr::filter(n_points > 0)

cat("Cells used:", nrow(pop_valid),
    "| Points to generate:", sum(pop_valid$n_points), "\n")

set.seed(42)
points_list <- lapply(seq_len(nrow(pop_valid)), function(i) {
  n    <- pop_valid$n_points[i]
  bbox <- sf::st_bbox(pop_valid[i, ])
  data.frame(
    x = runif(n, bbox["xmin"], bbox["xmax"]),
    y = runif(n, bbox["ymin"], bbox["ymax"])
  )
})

womenrepage <- sf::st_as_sf(
  do.call(rbind, points_list),
  coords = c("x", "y"),
  crs    = sf::st_crs(population_proj)
)

cat("Population points generated:", nrow(womenrepage), "\n")
```



## Step 4: Build spatial window from Kano polygon boundary

Using the actual Kano boundary polygon (not a bounding box rectangle)
ensures the risk surface matches the study area shape exactly.
This requires projected coordinates - hence the UTM reprojection above.


## Step 4 — Build observation window

Uses the actual buffered boundary polygon (not a bounding box).

```{r build-window, include=FALSE}
kano_window <- spatstat.geom::as.owin(sf::st_cast(kano_buffer, "POLYGON"))
```


## Step 5: Create PPP objects

PPP objects are two
- Cases: Maternal dates within Kano study area
- Controls: population at risk defined as women of childbearing age

```{r create-cases-ppp, include=FALSE}
# Cases: maternal deaths within Kano study area
deaths_coords <- sf::st_coordinates(maternal_deaths_kano)
deaths_ppp    <- spatstat.geom::ppp(
  x      = deaths_coords[, 1],
  y      = deaths_coords[, 2],
  window = kano_window
)
```

```{r create-controls-ppp, include=FALSE}
# Controls: population at risk (women of childbearing age)
pop_coords <- sf::st_coordinates(womenrepage)
pop_ppp    <- spatstat.geom::ppp(
  x      = pop_coords[, 1],
  y      = pop_coords[, 2],
  window = kano_window
)
```

## Step 6: Estimate risk surface

### Option A: Fixed bandwidth (run first to verify everything works)
```{r risk-fixed, include=TRUE}
rr_fixed <- risk(f = deaths_ppp, g = pop_ppp, doplot = TRUE)
```

### Option B: Geometric bandwidth (recommended)
```{r risk-geometric, include=TRUE}
h0 <- OS(deaths_ppp, nstar = "geometric")
cat("Optimal bandwidth:", h0, "\n")

rr_geom <- risk(
  f        = deaths_ppp,
  g        = pop_ppp,
  h0       = h0,
  tolerate = TRUE,
  doplot   = TRUE
)
```

### Option C: Adaptive bandwidth (most accurate, slowest)
# Set eval=TRUE in chunk header when ready to run - may take 10-30 mins
```{r risk-adaptive, eval=TRUE}
hpilot.f <- BOOT.density(deaths_ppp)
hpilot.g <- BOOT.density(pop_ppp)

rr_adapt <- risk(
  f               = deaths_ppp,
  g               = pop_ppp,
  h0              = h0,
  adapt           = TRUE,
  tolerate        = TRUE,
  hp              = c(hpilot.f, hpilot.g),
  pilot.symmetry  = "pooled",
  davies.baddeley = 0.05
)

plot(rr_adapt, main = "Risk Surface - Adaptive Bandwidth")
```

## Step 7 — Map risk surface (ggplot)

```{r plot-risk-surface}
# Convert to terra rasters and set CRS
rr_raster <- terra::rast(rr_adapt$rr); terra::crs(rr_raster) <- "ESRI:54009"
rr_p      <- terra::rast(rr_adapt$P);  terra::crs(rr_p)      <- "ESRI:54009"

# Crop and mask to study area
rr_raster <- terra::mask(terra::crop(rr_raster, terra::vect(kano_proj)), terra::vect(kano_proj))
rr_p      <- terra::mask(terra::crop(rr_p,      terra::vect(kano_proj)), terra::vect(kano_proj))

kano_sf <- sf::st_as_sf(kano_proj)

p_risk <- ggplot() +
  tidyterra::geom_spatraster(data = rr_raster) +
  scale_fill_gradientn(
    colours  = c("#000080", "#800080", "#FF0000", "#FF4500", "#FF8000", "#FFFF00"),
    na.value = "transparent",
    name     = "Log Risk\nRatio"
  ) +
  geom_sf(data = kano_sf, fill = NA, color = "black", linewidth = 0.5) +
  tidyterra::geom_spatraster_contour(
    data = rr_p, breaks = 0.05, color = "black", linewidth = 0.6
  ) +
  geom_sf(data = maternal_deaths_kano,
          color = "black", fill = "yellow", shape = 21, size = 1.5, stroke = 0.4) +
  ggspatial::annotation_scale(location = "bl", width_hint = 0.2, text_cex = 0.7) +
  ggspatial::annotation_north_arrow(
    location    = "br", which_north = "true",
    pad_x       = unit(0.1, "in"), pad_y = unit(0.1, "in"),
    style       = north_arrow_fancy_orienteering(text_size = 6)
  ) +
  labs(
    title    = "Maternal Death Risk Surface — Kano, Nigeria",
    subtitle = "Kernel density ratio (adaptive bandwidth) | Contours: p < 0.05"
  ) +
  theme_void(base_size = 11) +
  theme(
    plot.title    = element_text(face = "bold", size = 13, hjust = 0.5),
    plot.subtitle = element_text(size = 9, color = "grey40", hjust = 0.5),
    legend.position = "right"
  )

print(p_risk)
```
## Step 8 — Normalize risk surface

```{r normalize, include=FALSE}
# Min-max normalization to [0, 1]
rr_min  <- terra::global(rr_raster, "min", na.rm = TRUE)$min
rr_max  <- terra::global(rr_raster, "max", na.rm = TRUE)$max
rr_norm <- (rr_raster - rr_min) / (rr_max - rr_min)

cat("Normalized range: [",
    terra::global(rr_norm, "min", na.rm = TRUE)$min, ",",
    terra::global(rr_norm, "max", na.rm = TRUE)$max, "]\n")
```

## Step 9 — Extract risk at model grid centroids

Risk values are extracted at each model grid cell centroid.
This aligns the coarse risk raster (~500m) with the finer model grid,
avoiding resolution mismatch problems that arise from rasterizing in the
opposite direction.

```{r extract-risk-at-model, include=FALSE}
# Compute centroids of model grid polygons (Mollweide)
model_pts     <- sf::st_centroid(model_proj)

# Extract normalized risk value at each centroid
risk_at_model <- terra::extract(rr_norm, terra::vect(model_pts))[, 2]

cat("Model grid cells:", length(risk_at_model), "\n")
cat("Cells with valid risk:", sum(!is.na(risk_at_model)), "\n")
```


# Comparative Analysis

## Risk surface vs Accessibility Score — density comparison

Each variable is extracted independently: risk from the raster, accessibility from
the model grid attribute table. Log-transformation corrects right skew in accessibility.

```{r density-comparison}
# Risk: valid pixels from normalized raster
rr_vec <- as.vector(terra::values(rr_norm, na.rm = TRUE))

# Accessibility: directly from model attribute table, log-normalised
acc_valid <- model_proj$Accessibility_Index_Standard
acc_valid <- acc_valid[!is.na(acc_valid)]
acc_log   <- log(acc_valid + 1e-10)
acc_vec   <- (acc_log - min(acc_log)) / (max(acc_log) - min(acc_log))

df_combined <- data.frame(
  value  = c(rr_vec, acc_vec),
  source = rep(c("Risk Surface", "Accessibility Score"),
               c(length(rr_vec), length(acc_vec)))
)

neutral_val <- (0 - rr_min) / (rr_max - rr_min)

ggplot(df_combined, aes(x = value, fill = source)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = neutral_val, linetype = "dashed", color = "black") +
  annotate("text", x = neutral_val, y = Inf, vjust = 2, hjust = -0.1,
           label = "log RR = 0\n(no excess risk)", size = 3.2, color = "grey30") +
  scale_fill_manual(values = c("Risk Surface"        = "orange",
                               "Accessibility Score" = "steelblue")) +
  labs(
    title    = "Risk Surface vs Accessibility Score — Kano, Nigeria",
    subtitle = "Risk = normalized log RR | Accessibility log-transformed | Dashed line = no excess risk",
    x = "Normalized Value", y = "Density", fill = ""
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```


## Scatter plot — grid cell level correlation

```{r scatter-correlation}
vals <- data.frame(
  risk = risk_at_model,
  acc  = model_proj$Accessibility_Index_Standard
) |> na.omit()

r_val  <- round(cor(vals$risk, vals$acc, method = "spearman"), 3)
r2_val <- round(r_val^2, 3)

ggplot(vals, aes(x = risk, y = acc)) +
  geom_point(alpha = 0.3, size = 1.2, color = "#2E86AB") +
  geom_smooth(method = "lm", color = "#E84855", linewidth = 1, se = TRUE) +
  annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -1,
           label = paste0("ρ = ", r_val, "\nρ² = ", r2_val),
           size = 4.5, color = "grey20") +
  labs(
    title    = "Mortality Risk vs Accessibility — grid cell level",
    subtitle = "Each point = one model grid cell centroid | Spearman correlation",
    x = "Normalized Log Risk Ratio", y = "Accessibility Score"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title    = element_text(face = "bold"),
        plot.subtitle = element_text(color = "grey40"))
```


## Risk by deprivation category

```{r risk-by-deprivation}
df_dep <- data.frame(
  risk        = risk_at_model,
  deprivation = model_proj$result
) |>
  na.omit() |>
  dplyr::mutate(deprivation = factor(deprivation,
    levels = 0:2,
    labels = c("Low Deprivation", "Medium Deprivation", "High Deprivation")
  ))

cat("Cell counts:\n"); print(table(df_dep$deprivation))

kw <- kruskal.test(risk ~ deprivation, data = df_dep)
cat("\nKruskal-Wallis p =", kw$p.value, "\n")
cat("\nMedian risk by group:\n")
print(tapply(df_dep$risk, df_dep$deprivation, median))

ggplot(df_dep, aes(x = deprivation, y = risk, fill = deprivation)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.1, fill = "white", outlier.shape = NA) +
  scale_fill_manual(values = c(
    "Low Deprivation"    = "steelblue",
    "Medium Deprivation" = "orange",
    "High Deprivation"   = "#D6604D"
  )) +
  annotate("text", x = 2, y = 0.95,
           label = paste0("Kruskal-Wallis\np = ", format(kw$p.value, digits = 3)),
           size = 3.5) +
  labs(
    title = "Maternal Death Risk by Deprivation Category — Kano, Nigeria",
    x = "", y = "Normalized Risk Surface", fill = ""
  ) +
  theme_minimal(base_size = 11) +
  theme(plot.title      = element_text(face = "bold", size = 12),
        legend.position = "none")
```

### Pairwise Dunn test (Bonferroni correction)

```{r dunn-test}
dunn.test(df_dep$risk, df_dep$deprivation, method = "bonferroni")
```

---
# Spatial Autocorrelation

## Bivariate Local Moran's I — Risk × Accessibility

Analysis units are the risk raster pixels (~500m), which match the effective
spatial resolution of the kernel density estimate. Accessibility is aggregated
(mean) from the finer model grid onto each risk pixel. This ensures both
variables share the same spatial scale, avoiding artificial autocorrelation
from inserting identical risk values into many smaller grid cells.

```{r bivariate-lisa}
# Convert risk raster pixels to polygons
lisa_poly <- terra::as.polygons(rr_raster, dissolve = FALSE)
names(lisa_poly)[1] <- "risk"

# Aggregate accessibility — single mean at extract step only
acc_rast <- terra::rasterize(terra::vect(model_proj), rr_norm,
                             field = "Accessibility_Index_Standard")
lisa_poly$acc <- terra::extract(acc_rast, lisa_poly, fun = mean, na.rm = TRUE)[, 2]

# Convert to sf, drop NA
lisa_sf <- sf::st_as_sf(lisa_poly) |> na.omit()
cat("Valid pixels for LISA:", nrow(lisa_sf), "\n")

# Invert acc so high = high deprivation (aligns with risk direction)
lisa_sf$acc_inv <- 1 - lisa_sf$acc

# Queen contiguity weights
queen_w <- rgeoda::queen_weights(lisa_sf)

# Bivariate Local Moran's I: risk (x) vs inverted accessibility (y)
biv_lisa <- rgeoda::local_bimoran(queen_w, lisa_sf[, c("risk", "acc_inv")])

# Attach cluster labels
lisa_sf$cluster <- factor(
  rgeoda::lisa_clusters(biv_lisa, cutoff = 0.05),
  levels = c(1, 2, 3, 4, 0),
  labels = c("High-High", "Low-Low", "Low-High", "High-Low", "Not Significant")
)

cat("\nCluster counts:\n"); print(table(lisa_sf$cluster))

# Reproject to WGS84 for plotting
lisa_wgs84 <- sf::st_transform(lisa_sf, crs = 4326)

cluster_colors <- c(
  "High-High"       = "#D6604D",
  "Low-Low"         = "#4393C3",
  "Low-High"        = "#92C5DE",
  "High-Low"        = "#F4A582",
  "Not Significant" = "grey85"
)

ggplot() +
  geom_sf(data = lisa_wgs84, aes(fill = cluster), color = NA) +
  geom_sf(data = sf::st_transform(kano, crs = 4326),
          fill = NA, color = "black", linewidth = 0.5) +
  scale_fill_manual(values = cluster_colors, name = "Bivariate\nLISA") +
  ggspatial::annotation_scale(location = "bl", width_hint = 0.2) +
  ggspatial::annotation_north_arrow(
    location = "br",
    style    = north_arrow_fancy_orienteering(text_size = 6)
  ) +
  guides(fill = guide_legend(override.aes = list(color = NA))) +
  labs(
    title    = "Bivariate LISA — Maternal Death Risk vs Accessibility Deprivation",
    subtitle = "High-High = high risk + high deprivation (priority areas) | p < 0.05"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey40", size = 9),
    legend.key.size = unit(0.8, "cm")
  )
```

# Save Outputs

```{r save-outputs, include=FALSE}
dir.create("outputs", showWarnings = FALSE)

# Reproject risk raster to WGS84 and resample to model grid resolution
rr_wgs84 <- terra::project(rr_raster, "EPSG:4326")

res_x    <- mean(model_outputs$lon_max - model_outputs$lon_min)
res_y    <- mean(model_outputs$lat_max - model_outputs$lat_min)
template <- terra::rast(
  xmin       = min(model_outputs$lon_min),
  xmax       = max(model_outputs$lon_max),
  ymin       = min(model_outputs$lat_min),
  ymax       = max(model_outputs$lat_max),
  resolution = c(res_x, res_y),
  crs        = "EPSG:4326"
)

rr_resampled <- terra::resample(rr_wgs84, template, method = "bilinear")
cat("Saved raster resolution:", terra::res(rr_resampled), "\n")

# PNG map
png("outputs/risk-surface-kano.png", width = 1200, height = 900, res = 150)
print(p_risk)
dev.off()

# GeoTIFF (WGS84, model grid resolution)
terra::writeRaster(rr_resampled, "outputs/risk-surface-kano.tif", overwrite = TRUE)
```




