---
title: "Estimating Risk Surface"
output: html_notebook
---

# Preliminary tasks 

## Preparation
This notebook estimates a spatial risk surface for maternal deaths in Kano, Nigeria,
based on the kernel density ratio method: https://arxiv.org/pdf/1707.06888

Key optimizations applied:
- WorldPop raster is aggregated before point generation to reduce memory load
- Population points are capped at a maximum sample size (10,000)
- Fixed bandwidth used first for speed; adaptive bandwidth optional
- Duplicate chunk names fixed
- Cases/Controls variable naming corrected

### Install required packages
```{r install-packages, include=FALSE, eval=FALSE}
# Only run once - set eval=FALSE after first run
install.packages(c('sparr', 'terra', 'sf', 'spatstat', 'dplyr', 'ggplot2'))
```

### Load required packages
```{r load-packages, include=FALSE}
library(sf)
library(sparr)
library(terra)       # Modern replacement for raster package
library(spatstat)
library(dplyr)
library(ggplot2)
```

```{r}
install.packages("rlang")
```

---

# Load Data

## Model outputs
```{r read-model-outputs, include=FALSE}
data <- read.csv("data/model-output-acc-score.csv")
head(data)
```

## Maternal deaths (Cases)
```{r read-maternal-deaths, include=FALSE}
maternal_deaths <- sf::st_read("data/maternaldeaths.geojson", quiet = TRUE)
head(maternal_deaths)
```

## Study area boundary
```{r read-study-area, include=FALSE}
kano <- sf::st_read("data/grid-boundary-kano.gpkg", quiet = TRUE)
head(kano)
```

---

# Validation: Test with built-in PBC dataset

Before running on real data, validate the method using `sparr`'s built-in PBC dataset.

```{r pbc-setup}
data(pbc)
pbccas <- split(pbc)$case
pbccon <- split(pbc)$control
```

```{r pbc-basic-risk, include=FALSE}
# Basic risk surface with default bandwidth
risk(f = pbccas, g = pbccon, doplot = TRUE)
```

```{r pbc-adaptive-risk, include=FALSE}
h0 <- OS(pbc, nstar = "geometric")
pbc_rr <- risk(
  f = pbccas, g = pbccon,
  h0 = h0, adapt = TRUE, tolerate = TRUE,
  hp = OS(pbc) / 2,
  pilot.symmetry = "pooled",
  davies.baddeley = 0.05
)
plot(pbc_rr)
```



---

# Main Analysis: Maternal Death Risk Surface in Kano

## Step 1: Reproject all data to UTM Zone 32N
# 
# spatstat's as.owin() requires a projected coordinate system (metres),
# not geographic coordinates (degrees). WGS84 will cause as.owin() to fail.
# UTM Zone 32N (EPSG:32632) is appropriate for Kano, Nigeria.
# All datasets must use the same CRS for ppp objects to align correctly.

```{r reproject-data}
CRS_UTM <- 32632

kano_proj           <- sf::st_transform(kano, crs = CRS_UTM)
maternal_deaths_proj <- sf::st_transform(maternal_deaths, crs = CRS_UTM)

cat("All data reprojected to UTM Zone 32N\n")
```

## Step 2: Filter maternal deaths to Kano study area
#
# maternal_deaths contains records from the wider Kano region.
# Deaths outside the study area boundary have no corresponding 
# population Controls and would bias the risk surface.

```{r filter-deaths}
maternal_deaths_kano <- sf::st_filter(maternal_deaths_proj, kano_proj)
cat("Deaths within Kano study area:", nrow(maternal_deaths_kano),
    "out of", nrow(maternal_deaths_proj), "\n")

# Visual check
plot(sf::st_geometry(kano_proj), main = "Maternal Deaths within Kano Study Area")
plot(sf::st_geometry(maternal_deaths_kano), add = TRUE, 
     col = "red", pch = 16, cex = 1.2)
```

## Step 3: Generate population points from WorldPop raster
#
# The original raster_to_random_points() generates points equal to the 
# population value per cell (mean ~481, max ~8741), producing ~1.8M points
# and crashing R.
#
# Fix: scale down all values by SCALE_FACTOR before generating points.
# A cell with 500 women gets 5 points, a cell with 8741 gets 87 points.
# Relative proportions are preserved so the kernel density estimate is
# mathematically equivalent to the original approach.
# pmax(1, ...) ensures every cell contributes at least 1 point.
```{r generate-population-points, include=FALSE}

r <- terra::rast("data/Kano_Women_15-49_WorldPOP.tif")
r_df <- as.data.frame(r, xy = TRUE, na.rm = TRUE)
colnames(r_df) <- c("x", "y", "pop")

# Scale down population values proportionally.
# Original values (mean ~481, max ~8741) would generate ~1.8M points and crash R.
# Dividing by 100 reduces this to ~18,000 points while keeping relative densities identical.
# A cell with 500 women gets 5 points; a cell with 8741 women gets 87 points.
SCALE_FACTOR <- 100 #could try different scale to get optimal 
r_df$n_points <- pmax(1, round(r_df$pop / SCALE_FACTOR))
# pmax(1, ...) ensures every cell contributes at least 1 point so no population is lost

cat("Total points to generate:", sum(r_df$n_points), "\n")

# Get raster resolution to define cell boundaries for random point placement
res_x <- terra::res(r)[1]
res_y <- terra::res(r)[2]

set.seed(42) # For reproducibility
# For each cell, generate n_points random coordinates within the cell boundary.
# Points are scattered within the 1km x 1km cell rather than placed at the centre,
# which gives a more realistic spatial distribution.
points_list <- lapply(1:nrow(r_df), function(i) {
  n <- r_df$n_points[i]
  data.frame(
    x = runif(n, r_df$x[i] - res_x/2, r_df$x[i] + res_x/2),
    y = runif(n, r_df$y[i] - res_y/2, r_df$y[i] + res_y/2)
  )
})

# Combine all points into a single data frame and convert to sf object
points_df   <- do.call(rbind, points_list)
womenrepage <- sf::st_as_sf(points_df, coords = c("x", "y"),
                             crs = st_crs(maternal_deaths))

cat("Population points generated:", nrow(womenrepage), "\n")
```

## Step 4: Build spatial window from Kano polygon boundary
#
# Using the actual Kano boundary polygon (not a bounding box rectangle)
# ensures the risk surface matches the study area shape exactly.
# This requires projected coordinates - hence the UTM reprojection above.


```{r build-window}
kano_poly   <- sf::st_cast(kano_proj, "POLYGON")
kano_window <- spatstat.geom::as.owin(kano_poly)
class(kano_window)
```


## Step 5: Create PPP objects

```{r create-cases-ppp}
# Cases: maternal deaths within Kano study area
deaths_coords <- sf::st_coordinates(maternal_deaths_kano)
deaths_ppp <- spatstat.geom::ppp(
  x      = deaths_coords[, 1],
  y      = deaths_coords[, 2],
  window = kano_window
)
cat("Deaths PPP:\n")
summary(deaths_ppp)
```

```{r create-controls-ppp}
# Controls: population at risk (women of childbearing age)
womenrepage_proj <- sf::st_transform(womenrepage, crs = 32632)

pop_coords <- sf::st_coordinates(womenrepage_proj)
pop_ppp    <- spatstat.geom::ppp(
  x      = pop_coords[, 1],
  y      = pop_coords[, 2],
  window = kano_window
)
summary(pop_ppp)
```

## Step 6: Estimate risk surface

### Option A: Fixed bandwidth (run first to verify everything works)
```{r risk-fixed}
rr_fixed <- risk(f = deaths_ppp, g = pop_ppp, doplot = TRUE)
plot(rr_fixed, main = "Risk Surface - Fixed Bandwidth")
```

### Option B: Geometric bandwidth (recommended)
```{r risk-geometric}
h0 <- OS(deaths_ppp, nstar = "geometric")
cat("Optimal bandwidth:", h0, "\n")

rr_geom <- risk(
  f        = deaths_ppp,
  g        = pop_ppp,
  h0       = h0,
  tolerate = TRUE,
  doplot   = TRUE
)
plot(rr_geom, main = "Risk Surface - Geometric Bandwidth")
```

### Option C: Adaptive bandwidth (most accurate, slowest)
# Set eval=TRUE in chunk header when ready to run - may take 10-30 mins
```{r risk-adaptive, eval=FALSE}
hpilot.f <- BOOT.density(deaths_ppp)
hpilot.g <- BOOT.density(pop_ppp)

rr_adapt <- risk(
  f               = deaths_ppp,
  g               = pop_ppp,
  h0              = h0,
  adapt           = TRUE,
  tolerate        = TRUE,
  hp              = c(hpilot.f, hpilot.g),
  pilot.symmetry  = "pooled",
  davies.baddeley = 0.05
)
plot(rr_adapt, main = "Risk Surface - Adaptive Bandwidth")
```


## Step 7: Export results

```{r export}
png("outputs/risk-surface-kano.png", width = 1200, height = 900, res = 150)
plot(rr_adapt, main = "Maternal Death Risk Surface - Kano, Nigeria")
dev.off()
```








